---
title: "Data Viz 4 - Mapping Flows"
output: html_document
---

## Learning Objectives
By the end of this practical lab you will be able to:
* Manipulate LODES data from census blocks into an aggregate geography
* Format origin-destination data so that it is suitable for mapping
* Map flows using base plot and ggpolot

## Interaction data
The dynamics of urban systems are captured by various types of interaction data with varying degrees of spatio-temporal granularity. In this practical you will explore various ways in which interaction data can be mapped; including the estimation and representation of routes from co-ordinate pairs, mapping GPS trails and how very large origin-destination flows can be summarized.

## Estimating and representing routes
Many interaction data within cities are simply pairs of origin and destination locations, with some flow recorded between them. A good source of such data which are available within many municipalities relate to the location and flow between bike share docking stations. Many of the operators of these systems now make these data openly available.

We will now read in the September 2015 - August 2016 data from the [Bay Area Bike Share, SF, USA](http://www.bayareabikeshare.com/open-data). Although the original zip files contain other data, we will just read in the data related to the station location and also the use records.

```{r}
#Read in data
stations <- read.csv("./data/201608_station_data.csv")
trips <- read.csv("./data/201608_trip_data.csv")
```

Each of the stations has various attributes and cover a series of locations within the bay area - in this case, we will subset to only those within San Francisco.

```{r}
head(stations)
```

```{r}
# Limit to SF stations
stations <- stations[stations$landmark == "San Francisco",]
```


```{r warning=FALSE,message=FALSE}
library(ggmap)
```

```{r}
#Get background map for Chicago
SanFran <- get_map(location = c(-122.405,37.79), zoom = 14,color = 'bw')

#Basic point plot
ggmap(SanFran) + geom_point(data = stations, aes(x = long, y = lat, colour = "red")) + 
theme_bw() +
theme(axis.line = element_blank(),
      axis.text = element_blank(),
      axis.title=element_blank(),
      axis.ticks = element_blank(),
      legend.key = element_blank(),
      legend.position="none",
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank())
```

Turning now to the trips taken between these origin and destinations, these are ordered within the data frame as a trip per row, with each trip giving a various details including the start and end terminals. Using these data we will create a table of aggregate origin destination flows - however, we will only consider those flows between the stations identified as "San Francisco" and therefore shown on the map above.

```{r}
# View top six rows of the trips data
head(trips)
```

```{r}
# Get a list of the station IDs within SF
s_SF <- unique(stations$station_id)

# Limit trips to those with both origin and destination within the SF subset
trips_SF <- trips[(trips$Start.Terminal %in% s_SF) & (trips$End.Terminal %in% s_SF),]

# Create an table with origins and destination pairs
OD_trips_SF <- table(trips$Start.Terminal,trips$End.Terminal)
#View the top six rows
head(OD_trips_SF)
```

If you remember from an earlier practical we can convert a table object into a data frame as follows which turns it from a wide to a narrow format:

```{r}
# Create a data frame of the OD pairs
OD_trips_SF_Narrow <- data.frame(OD_trips_SF)
# Create sensible column names
colnames(OD_trips_SF_Narrow) <- c("Origin","Destination","Trips")
```

We will now identify the top ten most frequently ridden origin-destination pairs

```{r}
# Sorts the trips in decending order
OD_trips_SF_Narrow <- OD_trips_SF_Narrow[order(OD_trips_SF_Narrow$Trips,decreasing = TRUE),]
# Get the top 10 trips
top10 <- OD_trips_SF_Narrow[1:10,]
top10
```

We will now add origin and destination latitude and longitude co-ordinates by merging with the stations data. First the origin locations:

```{r}
# Add origin co-ordinates
top10 <- merge(top10,stations, by.x="Origin",by.y="station_id", all.x=TRUE)
# Remove unwanted columns
top10 <- subset(top10, select=c("Origin","Destination","Trips","lat","long"))
# Change column names
colnames(top10) <- c("Origin","Destination","Trips","O_lat","O_long")
```

And then the destinations:

```{r}
# Add destination co-ordinates
top10 <- merge(top10,stations, by.x="Destination",by.y="station_id", all.x=TRUE)
# Remove unwanted columns
top10 <- subset(top10, select=c("Origin","Destination","Trips","O_lat","O_long","lat","long"))
# Change column names
colnames(top10) <- c("Origin","Destination","Trips","O_lat","O_long","D_lat","D_long")
```

One of the simplest ways of calculating a route is to use the Google maps API which is implemented in the googleway package.

```{r eval=FALSE}
# Install package
install.packages("googleway")
```
```{r}
# Load package
library(googleway)
```

```{r echo=FALSE}
key <- "AIzaSyD5YyiOC01UDHRJvpgPE41ar19wqa9ZXAc"
```

For this you will need to get a [Google maps API](https://developers.google.com/maps/documentation/directions/) key:

```{r eval = FALSE}
# Set your key
key <- "your_api_key"
```

We will then extract an origin destination pair from our top10 object, and then use the google_directions() function to generate a route - this is then converted to a set of lat lon waypoints using decode_pl():

```{r}
# Using the first origin/destination
x <- 1 # You can change this between 1 - 10 to view each of the routes
origin <- c(top10[x,"O_lat"],top10[x,"O_long"])
destination <- c(top10[x,"D_lat"],top10[x,"D_long"])

# get the directions from Google Maps API
res <- google_directions(origin = origin,destination = destination,key = key, mode= "bicycling")

# Convert the results to co-ordinates
df_polyline <- decode_pl(res$routes$overview_polyline$points)

# See the top six rows
head(df_polyline)
```

These can then be mapped with ggmap:

```{r}
ggmap(SanFran) +
  geom_path(aes(x = lon, y = lat), color = "red", size = 0.5, data = df_polyline, lineend = "round")

```

We can extend the above to run a conditional statement with the for() function which does something (in this case, what is in brackets) until a condition is satisfied. Here loop changes the value of x from 1 to the number of rows in the top10 object (i.e. 10), and for each change in x the code between the { and } is run. For loops are very helpful to run a block of code multiple times.

Because x is changed from 1-10 on each run, we can use this value in various helpful ways, firstly to select a particular row from the data frame top10, and second to act as an ID for each set of routes extracted.


```{r}

tmp <-  data.frame(lat = numeric(0), lon = numeric(0), ID = numeric(0), Trips = numeric(0))

for (x in 1:nrow(top10)) {

# Get origins and destinations
origin <- c(top10[x,"O_lat"],top10[x,"O_long"])
destination <- c(top10[x,"D_lat"],top10[x,"D_long"])
  
# get the directions from Google Maps API
res <- google_directions(origin = origin,destination = destination,key = key, mode= "bicycling")

# Convert the results to co-ordinates
df_polyline <- decode_pl(res$routes$overview_polyline$points)

# Add a route ID and Trips to the data frame
df_polyline$ID <- x
df_polyline$Trips <- top10[x,"Trips"]

# Append the results to the tmp object
tmp <- rbind(tmp,df_polyline)

}


```

We can now visualize this using the ID as a factor which shows each route as a separate color. 

```{r}
ggmap(SanFran) +
  geom_path(aes(x = lon, y = lat,color = as.factor(ID)), size = 0.5, data = tmp, lineend = "round")
```

To enable some more experimentation with the flow data visualization without having to generate all the potential routes yourself, we have run these already for all origin destination station pairs where the flow was greater than 0. We will load these now:

```{r echo = FALSE, eval = FALSE}

# Create a list of all flows greater than 0 and within SF
ALL <- OD_trips_SF_Narrow[(OD_trips_SF_Narrow$Origin %in% s_SF) & (OD_trips_SF_Narrow$Destination %in% s_SF),]
ALL <- ALL[ALL$Trips > 0,]
ALL <- ALL[ALL$Origin != ALL$Destination,]
ALL$ID <- 1:nrow(ALL)

# Add origin co-ordinates
ALL <- merge(ALL,stations, by.x="Origin",by.y="station_id", all.x=TRUE)
# Remove unwanted columns
ALL <- subset(ALL, select=c("Origin","Destination","Trips","lat","long"))
# Change column names
colnames(ALL) <- c("Origin","Destination","Trips","O_lat","O_long")


# Add destination co-ordinates
ALL <- merge(ALL,stations, by.x="Destination",by.y="station_id", all.x=TRUE)
# Remove unwanted columns
ALL <- subset(ALL, select=c("Origin","Destination","Trips","O_lat","O_long","lat","long"))
# Change column names
colnames(ALL) <- c("Origin","Destination","Trips","O_lat","O_long","D_lat","D_long")


tmp_all <-  data.frame(lat = numeric(0), lon = numeric(0), ID = numeric(0), Trips = numeric(0))

for (x in 1:nrow(ALL)) {

# Get origins and destinations
origin <- c(ALL[x,"O_lat"],ALL[x,"O_long"])
destination <- c(ALL[x,"D_lat"],ALL[x,"D_long"])
  
# get the directions from Google Maps API
res <- google_directions(origin = origin,destination = destination,key = key, mode= "bicycling")

# Convert the results to co-ordinates
df_polyline <- decode_pl(res$routes$overview_polyline$points)

# Add a route ID and Trips to the data frame
df_polyline$ID <- ALL[x,"ID"]
df_polyline$Trips <- ALL[x,"Trips"]

# Append the results to the tmp object
tmp_all <- rbind(tmp_all,df_polyline)

Sys.sleep(time = 1)
print(x)

}

All_Flows <- tmp_all

save(All_Flows,file="All_Flows.Rdata")
```


```{r}
# Load flows
load("./data/All_Flows.Rdata")
# Show the top six rows of the table
head(All_Flows)
```

We can now show these on a map - we use the group option within the aes to tell ggmap that these are id that separate the routes, otherwise the whole set of co-ordinates are interpreted as a single route. You can remove these and generate the plot again to see what happens. 

```{r}
ggmap(SanFran) +
  geom_path(aes(x = lon, y = lat, group = ID), data = All_Flows)
```

For those who know the topography of San Francisco will understand why the Google routes have been calculated to avoid the central area.

<iframe width="560" height="315" src="https://www.youtube.com/embed/31JgMAHVeg0" frameborder="0" allowfullscreen></iframe>

We can now use the trip information to adjust the plot - for example, to scale the routes by the flow volume. We add the size option, but also divide the flows by 1000 to make the line widths an acceptable size. Thicker lines represent greater flows. We have also added the location of the stations in red.

```{r}
ggmap(SanFran) +
  geom_path(aes(x = lon, y = lat, group = ID), data = All_Flows,size = All_Flows$Trips/1000) +
  geom_point(data=stations, aes(long, lat),colour="red")
```

There are a lot of other adjustment options, for example, we can darken the map and change the line color:

```{r}
ggmap(SanFran,darken = 0.8) +
  geom_path(aes(x = lon, y = lat, group = ID), data = All_Flows, size = All_Flows$Trips/1000, colour = "white") +
  geom_point(data=stations, aes(long, lat),colour="red")

```

Or color the lines by intensity of flow; plus, we have also added some labels for the station ID using geom_text():


```{r}
ggmap(SanFran,darken = 0.8) +
  geom_path(aes(x = lon, y = lat, group = ID,colour = All_Flows$Trips), data = All_Flows, size = All_Flows$Trips/1000) +
  scale_colour_gradient(low="#900C3F", high="#FFC300") +
  geom_point(data=stations, aes(long, lat),colour="red") +
  geom_text(data = stations,aes(x = long, y = lat, label = station_id), check_overlap = TRUE, colour="#FFFFFF",hjust=-0.6)

```

We will now clean up this plot by removing the unwanted elements and changing the title of the legend:

```{r}
ggmap(SanFran,darken = 0.8) +
  geom_path(aes(x = lon, y = lat, group = ID,colour = All_Flows$Trips), data = All_Flows, size = All_Flows$Trips/1000) +
  scale_colour_gradient(low="#900C3F", high="#FFC300",name="Trips") +
  geom_point(data=stations, aes(long, lat),colour="red") +
  geom_text(data = stations,aes(x = long, y = lat, label = station_id), check_overlap = TRUE, colour="#FFFFFF",hjust=-0.6) +
  theme (
		axis.text = element_blank (), 
		axis.title = element_blank (),
		axis.ticks = element_blank ()
		)
```


## Tracking data

In the previous example we estimated the routes that cyclists may have taken as part of a bike share scheme - in reality the true routes taken would be divergent from these estimated paths which are essentially an optimized shortest path (based on how google calculate these for cyclists), and as such are a set of co-ordinates that follow the road topology.

For some urban analytics applications we may have tracking data gathered by GPS. For this example we will use a sample of the [GeoLife data](https://www.microsoft.com/en-us/download/details.aspx?id=52367). This GPS trajectory dataset was collected by 182 users during a period of three years from April 2007 to August 2012, and was part of a Microsoft Research Asia project. We will import a sample of this data which are the records for a single user. There are multiple text files with the file extension ".pit", each of which relate to a different journey. The first six rows of these data can be discarded. 

In total there are 395 of these files which as you could imagine would take a very long time to import one by one. As such, we have written a small block of code that will import these for you and store each imported data frame in a list. This is an alternative to the loop that was presented in the last section.

```{r message=FALSE, warning=FALSE}
library(lubridate)
```

```{r}
# Create a list of the files to import
file_list <- list.files("./data/GeoLife", full=T)
count <- length(file_list)

# This function imports a file
file_con <- lapply(file_list, function(x){
  tab <- read.table(x, head=F, quote = "\"", skip = 6, sep = ",") # import file
  colnames(tab) <- c("lat","lon","zero","alt","days","date","time") # change the column headings
  tab <- subset(tab,select=c("lat","lon","date","time")) # discard unwanted columns
  tab$date <- ymd(tab$date) # Set the time date
  tab$time <- hms(tab$time) # Set the time format
  tab$ID <- strsplit(x,"/|\\.")[[1]][5] # Uses the file name as an ID
  return(tab)
})

# Run the function and rbind the data frames together
file_con_df <- do.call(rbind, file_con)
```

The imported files are very similar in structure to the combined routes we explored in the last section. However, here, each GPS trail is separated by an ID.

```{r}
head(file_con_df)
```

We will first create a simple map of the densest area of the activity:

```{r}
#Get background map 
beijing <- get_map(location = c(116.3244, 39.99202), zoom = 13,color = 'bw')
#Create plot
ggmap(beijing) +
  geom_path(aes(x = lon, y = lat, group = ID), data = file_con_df)
```

With GPS gathered data the tracks can be messy and will not necessarily snap to network features such as paths or road. One way to make a GPS map more readily interpretable we can adjust the alpha of the plot.

```{r}
ggmap(beijing) +
  geom_path(aes(x = lon, y = lat, group = ID), data = file_con_df,alpha=0.1)
```

We can make this clearer by adjusting some of the colors:


```{r}
ggmap(beijing,darken = 0.8) +
  geom_path(aes(x = lon, y = lat, group = ID), data = file_con_df,alpha=0.1,colour="#FFFFFF") 
```

We can also use the wday function to split the gps data into weekdays or weekend and see if there are different patterns visible:

```{r}
ggmap(beijing,darken = 0.8) +
  # plots the weekday
  geom_path(aes(x = lon, y = lat, group = ID), data = file_con_df[!wday(file_con_df$date) %in% c(0,7),],alpha=0.1,colour="green") +
  # plots weekend
  geom_path(aes(x = lon, y = lat, group = ID), data = file_con_df[wday(file_con_df$date) %in% c(0,7),],alpha=0.1,colour="white")
```

We can also use a facet grid to visualize the patterns by a temporal period - in this case by hour:

```{r fig.width=7, fig.height=20}
# Create a new hour of the day variable
file_con_df$hour <- hour(file_con_df$time)

# Facet plot
ggmap(beijing,darken = 0.8, legend='none') +
  geom_path(aes(x = lon, y = lat, group = ID), data = file_con_df,alpha=0.2,colour="white") +
  facet_wrap(~hour,ncol=3) +
  theme (
		axis.text = element_blank (), 
		axis.title = element_blank (),
		axis.ticks = element_blank ()
		)
```

## Interaction data

At more disaggregate temporal scales flow data are captured within many urban contexts through more traditional survey data or censuses. These commonly are created for questions around home-work locations or migration; and as such for a variety of variables it is often possible to map aggregate flows both internally and within urban systems.

In this final section of the practical we will map some [longitudinal employer-household dynamics](https://lehd.ces.census.gov/data/) data from the US Census Bureau. There are a range of different data within this collection, but for these purposes we will utilize an extract of the origin destination data - LEHD Origin-Destination Employment Statistics (LODES). These are available for each state, however, the extract we consider here is for 2014 in Texas. This is at a block level.

First we will read in the origin-destination data and a cross walk file which gives a list of higher aggregations for the blocks. We have cut down both tables from the full file to save space.

```{r}
# Import OD
load("./data/tx_od_main_JT00_2014.Rdata")
# Import cross walk
XWalk <- read.csv("./data/tx_xwalk.csv")
# We will also turn off scientific notation for this example as the block codes are stored as a very large number
options(scipen=999)
```

```{r}
# View top six rows of OD data
head(OD)
```

There are a lot of variables within the file, the details of which can be found [here](https://lehd.ces.census.gov/data/lodes/LODES7/LODESTechDoc7.2.pdf); however, for this practical we will just be using "S000" which relates to the total number of jobs alongside the origin (homes - h_geocode) and destination (work - w_geocode) block codes. 

We will view the crosswalk data frame and then we need to append this to the OD data. We merge this twice as the tract codes are needed for both the origin and destination blocks.

```{r}
# View the top six rows of data
head(XWalk)

# Merge onto the home block code
OD <- merge(OD,XWalk, by.x = "h_geocode", by.y = "tabblk2010", all.x= TRUE)
# Change column names
colnames(OD) <- c("h_geocode","w_geocode","S000","h_geocode_trct")

# Merge onto the work block code
OD <- merge(OD,XWalk, by.x = "w_geocode", by.y = "tabblk2010", all.x= TRUE)
# Change column names
colnames(OD) <- c("h_geocode","w_geocode","S000","h_geocode_trct","w_geocode_trct")

```

We can then aggregate the flows into the origin and destination tracts:

```{r}
# Aggregate flows into Tracts
OD_Tract <- aggregate(data=OD, S000 ~ h_geocode_trct + w_geocode_trct, sum)
```

Before we can plot the flows we need the location of the tracts, which we can extract from a shapefile of the zone locations:

```{r message=FALSE, error=FALSE}
library(rgdal)
```

We will then read in the tract polygons for Texas downloaded from the [US Census Beurueau site](https://www.census.gov/geo/maps-data/data/cbf/cbf_tracts.html):

```{r message=FALSE, error=FALSE}
# Import spatial data
TX_SP <- readOGR("./data/Texas_Tract.geojson", "OGRGeoJSON")
# Convert to WGS84
TX_SP <- spTransform(TX_SP, CRS("+init=epsg:4326"))

```

We will now have a look at the content of the data frame associated with TX_SP and then use the coordinates() function which extracts the centroid of a zone to build a new lookup:

```{r}
# View the top 6 rows
head(TX_SP@data)

# Create lookup table
TX_tract_centroids <- data.frame(TX_SP@data$GEOID,coordinates(TX_SP))
# Change column names
colnames(TX_tract_centroids) <- c("Tract","lon","lat")
# View the top six rows of the new data
head(TX_tract_centroids)
```

We will now merge the tract coordinates onto the tract flow data frame - again, we will do this twice and adjust the column names so both origin and destinations are coded.

```{r}
# Add home lat lon
OD_Tract <- merge(OD_Tract,TX_tract_centroids,by.x="h_geocode_trct", by.y="Tract",all.x=TRUE)
# Fix column names
colnames(OD_Tract) <- c("h_geocode_trct","w_geocode_trct","S000","h_lon","h_lat")

# Add work lat lon
OD_Tract <- merge(OD_Tract,TX_tract_centroids,by.x="w_geocode_trct", by.y="Tract",all.x=TRUE)
# Fix column names
colnames(OD_Tract) <- c("w_geocode_trct","h_geocode_trct","S000","h_lon","h_lat","w_lon","w_lat")
```

We can now create a map - we will first create a map connecting tracts with flows of over 70

```{r message=FALSE, error=FALSE}
# Get base map
texas <- get_map(location = "Texas", zoom = 6,color = 'bw')
```

```{r}
# Create plot
ggmap(texas) + 
      geom_segment(data=OD_Tract[OD_Tract$S000 > 70,],aes(y = h_lat, x = h_lon, yend = w_lat, xend = w_lon))
```

This is a bit messy, but we can improve this plot by adjusting some colors, line widths and transparency:

```{r}
# Create plot
ggmap(texas,darken = 0.8) + 
      geom_segment(data=OD_Tract[OD_Tract$S000 > 70,],aes(y = h_lat, x = h_lon, yend = w_lat, xend = w_lon),colour= "white", alpha= 0.1, size=0.2)
```
We will now create a map at a city scale; in this case Houston. We have adjusted some of the parameters again, and also increased the number of origin-destinations:

```{r message=FALSE, error=FALSE}
# Get base map
Houston <- get_map(location = "Houston, TX", zoom = 10,color = 'bw')
```
```{r}
# Create plot
ggmap(Houston,darken = 0.8) + 
      geom_segment(data=OD_Tract[OD_Tract$S000 > 5,],aes(y = h_lat, x = h_lon, yend = w_lat, xend = w_lon),colour= "white", alpha= 0.01, size=0.2)

```

The problem with this plot is that all the origin destination pairs are being considered uniformly, and there is no consideration of the volume of flows. We can adjust this using the scale_alpha_continuous() option.

```{r}
# Create plot
ggmap(Houston,darken = 0.8) + 
      geom_segment(data=OD_Tract[OD_Tract$S000 > 5,],aes(y = h_lat, x = h_lon, yend = w_lat, xend = w_lon, alpha= S000), size=0.3, colour = "white") +
  scale_alpha_continuous(range = c(0.004, 0.3))  +
  theme ( legend.position="none",
		axis.text = element_blank (), 
		axis.title = element_blank (),
		axis.ticks = element_blank ()
		)
```

# Further resources / training

* [Mapping flows in R](http://spatial.ly/2015/03/mapping-flows/)