---
title: "7. Data Viz 3 - Visualizing Point Patterns"
output: html_document
---


##Learning Objectives
By the end of this practical lab you will be able to:
* Create 


- binning
- kde

We will first read some crime data into R for the [City of Chicago, USA](https://data.cityofchicago.org/view/5cd6-ry5g). This contains the recorded crimes during the year 2016 as of 24th December 2016; with the attributes including a category of the crime and a variety of other information such as location and date / time. The following code imports the data, parses the date and time stamp using the base R function strptime() and also ymd_hms() from the package lubridate. We then remove unwanted columns and restrict the results to "BURGLARY".


```{r}
# Import Crimes
crimes <- read.csv("./data/chicago_crimes_2016.csv")

# Append days, months
install.packages("lubridate")
library(lubridate)

#Parse date & time
crimes$New_Date <- ymd_hms((strptime(crimes$Date, "%m/%d/%Y %I:%M:%S %p",tz="UTC")))

#Subset the data to remove unwanted colums
crimes <- crimes[crimes$Primary.Type == "BURGLARY",c("ID","Latitude","Longitude","New_Date")]

#View the top of the data
head(crimes)
```

We can then see how the burgalaries are distributed by day of the week:

```{r}
ggplot(data=crimes, aes(wday(crimes$New_Date,label = TRUE))) +
    geom_bar() +
    xlab("Day") + 
    ylab("Burglaries (count)")
```

Or by month:

```{r}
ggplot(data=crimes, aes(month(crimes$New_Date,label = TRUE))) +
    geom_bar() +
    xlab("Month") + 
    ylab("Burglaries (count)")
```

We can also use the facet_grid option within ggplot to produce separate plots for each month; and additionally add an aesthetic to the geom_bar which colors each day differently.

```{r}
ggplot(data=crimes, aes(wday(crimes$New_Date,label = TRUE))) +
    geom_bar(aes(,fill=wday(crimes$New_Date,label = TRUE))) +
    xlab("Day") + 
    ylab("Burglaries (count)") +
    facet_grid(~month(crimes$New_Date,label = TRUE)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),legend.position="none")
```

Using a similar technique we can also explore the time of the day in which burglaries are recorded broken down into:

```{r}
#Create a summary data frame of the counts of burglaries by hour time band and quarter of the year
t <- data.frame(table(hour(crimes$New_Date),quarter(crimes$New_Date)))
colnames(t) <- c("Time","Quarter","Freq") # Name columns as something sensible

#Create plot
p <- ggplot(t, aes(x=Time, y=Freq, group=Quarter, colour = Quarter))
p + geom_line()

```







We will now consider 

```{r}


library(ggmap)
#Plot
chicago <- get_map(location = "chicago", zoom = 11)
ggmap(chicago) + geom_point(data = crimes, aes(x = Longitude, y = Latitude))
```

```{r eval=FALSE}
install.packages("pointdensityP")
```
```{r}
library(pointdensityP)
```




```{r}
# Remove crimes with no lat / lon
crimes <- crimes[!is.na(crimes$Latitude),]

chicago <- get_map(location = "chicago", zoom = 11)


ggmap(chicago, base_layer = ggplot(crimes, aes(x=Longitude, y=Latitude))) +
coord_cartesian(xlim = c(-87.84918,-87.3)) +
stat_binhex(bins=20) + 
theme_bw() +
theme(axis.line = element_blank(),
      axis.text = element_blank(),
      axis.title=element_blank(),
      axis.ticks = element_blank(),
      legend.key = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank())



```


```{r}


ggmap(chicago, base_layer = ggplot(crimes, aes(x=Longitude, y=Latitude))) +
stat_bin2d(bins = 20)
```


```{r}


  
ggmap(chicago, base_layer = ggplot(crimes, aes(x=Longitude, y=Latitude))) +
  stat_density2d(aes(x = Longitude, y = Latitude,fill = ..level..,alpha=..level..), bins = 10, geom = "polygon", data = crimes) +
  scale_fill_gradient(low = "black", high = "red")


```



```{r}
# Calculate Density
density <- pointdensity(crimes,lat_col="Latitude",lon_col="Longitude",grid_size=1,radius=3)



map_base <- qmap('Chicago', zoom = 11, darken=0.2) 
map_base + geom_point(aes(x = lon, y = lat, colour = count,alpha=count), 
    shape = 16, size = 1, data =density) + 
    scale_colour_gradient(low = "yellow", high = "red") 
```

